A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950

### 2022-01-07 paper.Rmd
as usual

```{r figgraph, fig.cap = "The log-cube transform g."}
##g <- function(x) (log(1 + x))^3  # Note: days!
x <- seq(0, 365, length = 1000)
y <- g(x)
plot(x, y, type = "l", xlab = "age (days)", ylab = "g(age)", axes = FALSE, col = "blue")
xat <- c(0, 28, 365)
yat <- g(xat)
axis(1, at = xat)
axis(2, at = yat, labels = round(yat, 1), las = 1)
box()
lines(c(30, 30), c(0, g(30)), lty = 2, col = "darkgreen")
lines(c(0, 30), c(g(30), g(30)), lty = 2, col = "darkgreen")
abline(h = 0, v = 0)
abline(a = 0, b = 1, col = "red")
```

@@

From the Umeå-Skellefteå region, we follow all persons born in the 
region 1895--1950 until death, reaching the age of 1 year, or leaving follow-up for whatever 
reason, whichever comes first. The first few lines (infants) are

@@

We plot without the first step in the top panel and with it in the
bottom panel, see Figure \@ref(fig:ggplot2).

```{r ggplot2, fig.cap = "BB plots with and without time transform.", fig.height = 5}
nbths <- NROW(infdat)
##oldpar <- par(mfrow = c(2, 1))
sur <- with(infdat, Surv(enter, exit, event))
rs <- risksets(sur)
ss <- cumsum(rs$n.events)
## Natural:
xt <- rs$risktimes
att <- c(0, 28, 183, 365)
plot(xt, ss / nbths, ylim = c(0, max(ss / nbths)), type = "l",
     axes = FALSE, col = "blue", lwd = 2, xlab = "Age (days)",
     ylab = "Cum. deaths")
abline(h = 0)
axis(2, las = 1)
axis(1, at = att, labels = att)
box()
## g-transformed:
lines(g(xt) / g(365) * 365, ss / nbths, ylim = c(0, max(ss / nbths)), type = "l",
     col = "red", lwd = 2, xlab = "Age (days)",
     main = "After g-transform", ylab = "Cum. deaths")
##abline(h = 0)
##axis(2, las = 1)
axis(3, at = g(att) / g(365) * 365, labels = att, col = "red")
box()
abline(v = 0)
abline(v = g(c(28, 183, 365)) / g(365) * 365, col = "red", lty = 3)
abline(v = c(28, 183, 365), col = "blue", lty = 3)
##
legend("bottomright", legend = c("Natural", "g-transform"), col = c("blue", "red"), lty = 1)
abline(h = (ss / nbths) [28], lty = 3 )
abline(h = (ss / nbths) [180], lty = 3 )
##par(oldpar)
```

>>>>

```{r figgraph, fig.cap = "The log-cube transform g."}
##g <- function(x) (log(1 + x))^3  # Note: days!
x <- seq(0, 365, length = 1000)
y <- g(x)
plot(x, y, type = "l", xlab = "age (days)", ylab = "g(age)", axes = FALSE, col = "blue")
xat <- c(0, 28, 183, 365)
yat <- g(xat)
axis(1, at = xat)
axis(2, at = yat, labels = round(yat, 1), las = 1)
box()
lines(c(28, 28), c(0, g(28)), lty = 2, col = "darkgreen")
lines(c(0, 28), c(g(28), g(28)), lty = 2, col = "darkgreen")
##
lines(c(183, 183), c(0, g(183)), lty = 2, col = "darkgreen")
lines(c(0, 183), c(g(183), g(183)), lty = 2, col = "darkgreen")
##
lines(c(365, 365), c(0, g(365)), lty = 2, col = "darkgreen")
lines(c(0, 365), c(g(365), g(365)), lty = 2, col = "darkgreen")
abline(h = 0, v = 0)
##abline(a = 0, b = 1, col = "red")
```

@@

From the Umeå-Skellefteå region, we follow all persons born in the 
region 1895--1950 until death, reaching the age of 1 year, or leaving follow-up for whatever 
reason, whichever comes first. The first few lines (infants) are shown in Table 
\@ref(tab:infdat3).

@@

The result is shown in Figure \@ref(fig:ggplot2), with the original, untransformed
data together with the transform, stretched to 365.

```{r ggplot2, fig.cap = "BB plots with and without time transform, Umeå-Skellefteå 1895-1950.", fig.height = 5}
nbths <- NROW(infdat)
##oldpar <- par(mfrow = c(2, 1))
sur <- with(infdat, Surv(enter, exit, event))
rs <- risksets(sur)
ss <- cumsum(rs$n.events)
## Natural:
xt <- rs$risktimes
att <- c(0, 28, 183, 365)
plot(xt, ss / nbths, ylim = c(0, max(ss / nbths)), type = "l",
     axes = FALSE, col = "blue", lwd = 2, xlab = "Age (days)",
     ylab = "Cum. deaths")
abline(h = 0)
axis(2, las = 1)
axis(1, at = att, labels = att)
box()
## g-transformed:
lines(g(xt) / g(365) * 365, ss / nbths, ylim = c(0, max(ss / nbths)), type = "l",
     col = "red", lwd = 2, xlab = "Age (days)",
     main = "After g-transform", ylab = "Cum. deaths")
##abline(h = 0)
##axis(2, las = 1)
axis(3, at = g(att) / g(365) * 365, labels = att, col = "red")
box()
abline(v = 0)
abline(v = g(c(28, 183, 365)) / g(365) * 365, col = "red", lty = 3)
abline(v = c(28, 183, 365), col = "blue", lty = 3)
##
legend("bottomright", legend = c("Natural", "g-transform"), col = c("blue", "red"), lty = 1)
abline(h = (ss / nbths) [28], lty = 3 )
abline(h = (ss / nbths) [180], lty = 3 )
##par(oldpar)
```

The plot, disregarding the transform, is just a graph of the empirical *cumulative distribution function (CDF)*,
that is, one minus the *Kaplan-Meier estimate* of the *survivor function*.

## Post-neonatal mortality

We begin by studying mortality after age 28 days on the g-scale, BP predicts that it is approximately *uniform*. 
The uniform distribution, however, is not very useful in a survival analysis context, so we will
investigate a close alternative, the *exponential * distribution. Modifying the BP plot by using successive 
divison by *risk set size* instead of sample size should result in an approximately linear graph if the 
exponential assumption is valid.

```{r pneo, fig.cap = "Relation between cumulative hazards and CDF for transformed post-neonatal mortality data."}
ginfdat <- age.window(infdat, c(28, 365))
ginfdat$enter <- g(ginfdat$enter)
ginfdat$exit <- g(ginfdat$exit)
grs <- risksets(with(ginfdat, Surv(enter, exit, event)), members = FALSE)
fit <- coxreg(Surv(enter, exit, event) ~ 1, data = ginfdat)
ser <- plot(fit, fn = "sur", fig = FALSE)
##plot(fit, col = "blue", axes = FALSE)
cum <- fit$hazards[[1]]
x <- cum[, 1]
ch <- cumsum(cum[, 2])
plot(x, ch, type = "s", axes = FALSE, col = "blue", xlab = "Age (days)", ylab = "")
axis(1, at = c(g(28), g(183), g(365)), labels = c(28, 183, 365))
axis(2, las = 1)
abline(h = 0)
box()
x <- ser[[1]][[1]][, 1]
y <- ser[[1]][[1]][, 2]
lines(x, 1 - y, col = "red", lty = 2)
legend("topleft", legend = c("Cumulative hazards", "CDF (Bourgeouis-Pichat)"), col = c("blue", "red"),
       lty = c(1, 2))
```

As is evident from Figure \@ref(fig:pneo), the difference between the the two curves is
negligible, and both are well approximated by a straight line, so we will from now on work with the 
assumption that the transformed post-neonatal data are *exponentially distributed*.
There is also a mathematical argument motivating the similarity: The cdf for an exponentially 
distributed random variable $T$ is

\begin{equation*}
F_T(t) = 1 - \exp(-\lambda t), \quad t, \lambda > 0, 
\end{equation*}

and the first order approximation of this cdf is

\begin{equation*}
F_T(t) \approx \lambda t, \quad t, \lambda > 0,
\end{equation*}

which is the *cumulative hazards* function of $T$,
and the approximation is good for small values of $\lambda t$.
